Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-10 draft.tex
added some minor changes of the reviewer. Added changelog to keep track for cover letter

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
\\\\
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
%
The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
\\\\
Our contributions are:
\begin{itemize}
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them. The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

@@

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.

Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

Indicators for medical conditions as well as questions about a subject's lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.

@@

\section{Prior and Related Work}
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

\paragraph{Visualizing Image and Non-Image Data. }
Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space \cite{Blaas2007}.
%
They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.
%
Steenwijk and colleagues employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
They are useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
%
Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender \cite{Chui2011}.

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
The distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
%
User specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
%Anatomic covariation induced by user input is displayed using tensors 
%Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

% *********************** Our own Work ***********************
% Prior Work
\paragraph{SHIP Data Analysis.}
Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
%
Niemann and colleagues presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease \cite{Niemann2014}.
%
Association rules created by data mining methods can be analyzed interactively with their presented tool and highlight potentially overlooked features.
% *********************** / Our own Work ***********************

\paragraph{Interactive Visual Analysis}
% Interactive Visual Analysis - Steffens Paper
The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation \cite{Oeltze2007}.
%
This was focused on individual patients and not on cohort study data. 
%

@@

%
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
%
Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes.
%%TODO Adaptive visualization!
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
%
Each plot can be brushed using widgets.
%
Brush selections are propagated to all visualizations allowing for fast feature querying.
% TODO Timeline
%
%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
%
%A use case for this is when a continuous feature has to be divided into even groups.

>>>>

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
% TODO: Self added Packages - need to be removed afterwards
%\usepackage{babel}
\usepackage[normalem]{ulem}
\usepackage{color}
\newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people who are randomly selected to avoid any bias.
%
\rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
\add{The people are selected without focus on a certain disease.}
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
%
This also applies to the chosen landmarks used to quantify medical image data information.
%
The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
%
Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
\\\\
We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
%
Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
%
\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
\\\\
Our contributions are:
\begin{itemize}
	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
%	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
\end{itemize}
%
\rem{The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists}.
% http://wiki.infowiss.net/Gulf_of_execution ?
We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
%
This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
%
\add{We implement the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
%
This aims to trigger a fast feedback loop between us and the epidemiologists.}

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them. The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

@@

To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
%
The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
%
\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
%
%An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
A famous example is the Berkeley University gender bias case, where there were significantly less women admitted then man.
%
Analysis later showed that women applied mostly for strongly restricted subjects and when only looking acceptance divided by departments there was actually a small bias toward women.
%
%For example, insulin dependent and non-insulin dependent diabetes shows 
%In a famous example, researchers concluded that a newer treatment for kidney stones was more effective than traditional surgery, but it was later revealed that the newer treatment was more often being used on small kidney stones.
%For example, comparing mortality rate of non-insulin dependent 
%When for example surgeons are ranked regarding diseased patients, without including the risk of the surgeries, the results may be distorted.
}

Statistical tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

Indicators for medical conditions as well as questions about a subject's lifestyle are \rem{also} often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%
Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is \rem{kept, if possible, on the initial software and hardware version} \add{not updated or changed} during a longitudinal study to ensure comparability in and between acquisition cycles.

@@

\section{Prior and Related Work}
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
%Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
\rem{Their} \add{Turkay and colleagues \cite{Turkay2013}} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes \rem{the work of Turkay and colleagues} \add{their work} closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \rem{\cite{Turkay2013}}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

\paragraph{Visualizing Image and Non-Image Data. }
Gresh \rem{and colleagues} \add{et al. \cite{Gresh2000}} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
%
Blaas \rem{and colleagues} \add{et al. \cite{Blaas2007}} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
%
They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.
%
Steenwijk \rem{and colleagues} \add{et al. \cite{Steenwijk2010}} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
%
Zhang \rem{and colleagues} \add{et al. \cite{Zhang2012}} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
They are useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai \rem{and colleagues} \add{et al. \cite{Dai2005}} explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
%
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
%
Chui \rem{and colleagues} \add{et al. \cite{Chui2011}} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban \rem{and colleagues} \add{et al. \cite{Caban2011}} investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking \rem{and colleagues} \add{et al. \cite{Busking2010a}} in a 2D-projected plane of the space.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
The distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann \rem{and colleagues} \add{et al. \cite{Hermann2014}} identify local deformation changes by investigating shape-related differences on rodent mandibles.
%
User specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
%Anatomic covariation induced by user input is displayed using tensors 
%Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

% *********************** Our own Work ***********************
% Prior Work
\paragraph{SHIP Data Analysis.}
Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their presented tool and highlight potentially overlooked features.
% *********************** / Our own Work ***********************

\paragraph{Interactive Visual Analysis}
% Interactive Visual Analysis - Steffens Paper
The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
%
Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation.
%
This was focused on individual patients and not on cohort study data. 
%

@@

%
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, \add{information derived from the} medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
%
Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes.
%%TODO Adaptive visualization!
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
%
Each plot can be brushed using widgets.
%
Brush selections are propagated to all visualizations allowing for fast feature querying.
% TODO Timeline
%
%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
%
%A use case for this is when a continuous feature has to be divided into even groups.
